CITK - an architecture and examples of CUDA enabled ITK filters

### 2011-05-11 article/Article.tex
minor article update
renamed patch file

\begin{abstract}
\noindent
There is great interest in the use of graphics processing units (GPU)
for general purpose applications because the highly parallel
architectures used in GPUs offer the potential for huge performance
increases. The use of GPUs in image analysis applications has been
under investigation for a number of years. This article describes
modifications to the InsightToolkit (ITK) that provide a simple
architecture for transparent use of GPU enabled filters and examples
of how to write GPU enabled filters using the NVIDIA CUDA tools.
\end{abstract}

@@

The approach used in CITK does require a modification to core ITK
classes, but has a number of advantages. The same approach has since
been outlined on the ITK Wiki.

The fundamental component of the pipeline is the ITK Image
class. Within this class is a pixel container called {\em
  ImportImageContainer}, used to manage the image data. CITK includes
a substitute pixel container named {\em
  CudaImportImageContainer}. This pixel container has all the same
functionality of the {\em ImportImageContainer} which results in full
compatibility with existing ITK components.

The {\em CudaImportImageContainer} not only managed the image data on
the host, but it also managed the image data on the device. When a
standard filter requested the image data, such as through an iterator,
the {\em CudaImportImageContainer} checks whether the most up to date
image is on the device or the host. If it is on the device, it is
copied back onto the host. This data is then supplied to the
user. Similarly when a GPU filter requests the image data, the {\em
  CudaImportImageContainer} would check where the most up to date
image is, and copy it to the device if required.

@@

@@

Your CUDA kernel function should return a pointer to the output image as a global device pointer:

{\tt typename TOutputImage::PixelType * ptr;}

A pointer to input image memory is retrieved using:

{\tt input->GetDevicePointer();}

After the filter has completed, the output pointer must be given to
the output image to be passed down the pipeline:

{\tt output->GetPixelContainer()->SetDevicePointer(ptr, N, true);}

In this example the {\tt ptr} is the pointer to output memory and the 'N'
is the number of pixels in this image. The 'true' value is very
important. This tells the output image to free the device memory once
it is finished with it. WARNING: Setting this to false can lead to
memory leaks on the device.

>>>>

\begin{abstract}
\noindent
There is great interest in the use of graphics processing units (GPU)
for general purpose applications because the highly parallel
architectures used in GPUs offer the potential for huge performance
increases. The use of GPUs in image analysis applications has been
under investigation for a number of years. This article describes
modifications to the InsightToolkit (ITK) that provide a simple
architecture for transparent use of GPU enabled filters and examples
of how to write GPU enabled filters using the NVIDIA CUDA tools.

This work was performed between late 2009 and early 2010 and is being
published as modifications to ITK 3.20. It is hoped that publication will
help inform more general development of GPU support in ITK 4.0 and
facilitate experimentation by users requiring functionality of 3.20 or
wishing to pursue CUDA based developments.
\end{abstract}

@@

The approach used in CITK does require a modification to core ITK
classes, but has a number of advantages. A similar approach has since
been outlined on the ITK Wiki {\url http://www.cmake.org/Wiki/ITK_Release_4/GPU_Acceleration}.

The fundamental component of the pipeline is the {\em itk::Image}
class. Within this class is a pixel container called {\em
  ImportImageContainer}, used to manage the image data. CITK includes
a substitute pixel container named {\em
  CudaImportImageContainer}. This pixel container has all the same
functionality of the {\em ImportImageContainer} which results in full
compatibility with existing ITK components.

The {\em CudaImportImageContainer} manages image data on both the host
and device. When a standard filter requests the image data, such as
through an iterator, the {\em CudaImportImageContainer} checks whether
the most up to date image is on the device or the host. If it is on
the device, it is copied back onto the host. This data is then
supplied to the user. Similarly when a GPU filter requests the image
data, the {\em CudaImportImageContainer} would check where the most up
to date image is, and copy it to the device if required.

@@

\subsection{Weaknesses}
\begin{itemize}
\item This framework only supports Cuda, and not OpenCL. The ITK 4.0
proposal supports OpenCL. Limitations of the Cuda development
environment mean that even Cuda integration is less complete than
hoped, with significant changes to compilation processes being
necessary (see below).
\item A copy between host and device always results in the source of
the copy being considered redundant. This could be inefficient in some
cases. The problem is correctly dealing with identical copies on both
host and device. If, for example, a pipeline is branched such that one
branch in on GPU and the other on CPU, then the branch point is likely
to become a source of redundant copies.
\item The need to copy between device and host memory breaks some of
the usual assumptions, leading to some ugly use of {\em mutable}
declarations in CudaImportImageContainer.
\item Morphology filters included in the package use texture memory
but are not as generic as the standard ITK versions.
\end{itemize}

\section{Installation and building}
\subsection{Cuda compiler and software development kit}
This framework requires Cuda 3.2 and the SDK.
\subsection{Patch ITK 3.20}
The code distributed with this article includes a patch to modify ITK 3.20. This can be applied as follows:

\begin{itemize}
\item {\url http://voxel.dl.sourceforge.net/sourceforge/itk/InsightToolkit-3.20.0.tar.gz}
\item extract
\item cd ITK-3.20
\item patch -p0 < path/to/patch_3.20.dif
\end{itemize}

Alternatviely, this code may currently be retrieved via git and gerrit, as follows:
\begin{itemize}
\item git clone git://itk.org/ITK.git
\item cd ITK
\item git checkout v3.20.0
\item git pull http://review.source.kitware.com/p/ITK refs/changes/52/1452/1
\end{itemize}

\subsection{Build and install modified ITK}
There are many options available when building ITK. This process has
been tested under Linux and there are a number of changes to defaults
required to avoid limitations to the cuda development tools. 
\begin{itemize}
  \item Specify location of Cuda SDK.
  \item Turn off SSE options for VNL - advanced/VNL. This avoids errors caused by multiple inclusion of SSE files.
\end{itemize}

\subsection{Build examples}
\subsection{Workarounds for nvcc weaknesses/bugs}
\begin{itemize}
  \item multiple include of SSE files - turn off SSE options for VNL - advanced/VNL.
\end{itemize}

\subsection{Changes to standard processes for building ITK applications}
Typical application development in ITK utilizes templates and generic
programming and therefore does not require that the developer track
new object code dependencies when adding new filters. In principle the
same procedure should be possible when using CUDA enabled devices by
compiling all application code with nvcc, leading to non-cuda code
being compiled with the host c++ compiler and cuda code being compiled
with cuda compilers. This would also allow useful templating of cuda
kernels, leading to a relatively seamless integration with traditional
ITK development. Unfortunately the current generation of CUDA tools is
not able to cope with c++ of the complexity used in ITK. It is
therefore necessary to compile cuda kernels separately, which means
the developer must specify the correct object dependencies to the
linker. Examples of this can be seen in the CMake files included with
this article.

Alternatives, such as compiling all cuda kernels into a library, are
feasible but haven't been tested during this development.

@@

Two base classes have been provided to handle standard filter memory
management - {\em CudaInPlaceImageFilter} and {\em
CudaImageToImageFilter}. These filters allow the standard allocation
structure to be used via {\em this->AllocateOutputs()}. Explicit
allocation of device memory can be achieved using
Image->AllocateGPU().

\subsection{Templated kernel files}
ITK filters are generic with respect to pixel type and dimension.
